Double Standards in Judging Collective Action

### 2022-06-16 Documents/manuscript.Rmd
update manuscript

---
title: "Double Standards in Judging Collective Action"
header: "DOUBLE STANDARDS"
keywords: protest, collective action, double standards, social identity theory, system justification theory, Black Lives Matter
abstract: Collective action is a powerful force driving social change but, at the same time, often sparks contention about what are acceptable forms of collective action.
bibliography: references.bib
csl: apa.csl
output:
  pdf_document:
    latex_engine: pdflatex
    keep_tex: yes
    template: preprint-template.tex

@@

In this paper, we examine double standards in judging collective action---that is, whether observers judge the same protest actions as more or less acceptable depending on their own and the protesters' group memberships (*identity-based double standards*) or on the protesters' cause and how it aligns with their own ideological positions (*ideology-based double standards*). In two studies, we first develop an instrument of 25 controversial protest actions, based in item response theory, to capture double standards in judging collective action. In two preregistered experiments, we then use this instrument to test for double standards in judging protest actions for workers' rights in the United Kingdom and for or against defunding the police in the United States. We find evidence for ideology-based, but not for identity-based, double standards: People judge the same protest actions to be more acceptable when the protesters' cause aligns with their own ideological positions---but show no ingroup bias when judging collective action by ingroup and outgroup members.

<!-- Below, ... we situate the distinction ... below we introduce the distinction between normative ... derive predictions about ... from ... and introduce the present research ... -->

@@

Collective action can take many forms. Psychologists distinguish between *normative* collective action that seeks to achieve a political goal while conforming to the norms of the existing social system and *non-normative* collective action that violates those norms [@wright_responding_1990]. Recent research has shown that this distinction matters for how people respond to collective action. @shuman_disrupting_2021 showed that non-normative, non-violent collective action is more effective than either normative or violent collective action at gaining concessions from those most resistant to social change. @feinberg_activists_2020 demonstrated, however, that observers are less supportive of social movements that use extreme protest actions. Similarly, @teixeira_is_2020 found that advantaged-group members perceive non-normative collective action by disadvantaged-group members as more damaging to their ingroup's social image and are thus less supportive of such action. This means that activists face the dilemma that non-normative collective action might be most effective at gaining concessions but that it also risks reducing popular support for a cause. Together, these studies show that the distinction between normative and non-normative collective action is psychologically and societally consequential.

@@

Just as system justification shapes support for, and opposition to, collective action, we propose that it results in ideology-based double standards in what actions people consider acceptable forms of collective action. First, we hypothesize that  people will judge the same protest actions as more acceptable if the protest supports, rather than challenges, the system (*Hypothesis 2a*). This hypothesis reflects the assumption that the motivation to justify the system is, to some extent, universal. Second, we hypothesize that people will judge the same protest actions as more acceptable if they endorse system-justifying beliefs and the protest supports the system or if they reject system-justifying beliefs and the protest challenges the system (*Hypothesis 2b*). This hypothesis implies an symmetry in how people with different ideological orientations judge system-supporting and system-challenging collective action.

@@

In this research, item response theory serves two purposes. First, item response theory provides a formal and statistical definition of double standards in judging collective action as respondents being more accepting of the same protest actions depending on who the protesters and what they are protesting. In Experiments 1 and 2, we operationalize double standards as the differences in $\theta_{(z)}$ between conditions which, as $\theta_{(z)}$ is $z$-standardized, correspond to Cohen's $d$ effect sizes. Second, item response theory provides a statistical framework for scale development. In Studies 1 and 2, we select the most discriminating and informative actions to build an instrument for capturing double standards in the experiments.

@@

Our research serves two purposes. The first purpose is to develop an instrument to measure where people draw the line between acceptable and unacceptable forms of collective action. In Study 1, we ask participants to generate more and less extreme protest actions and compile a pool of protest actions for scale development. In Study 2, we ask participants to rate how acceptable they consider each of those protest action to be and use item response theory to select the most discriminating and informative protest actions for our instrument. In so doing, we build an instrument that enables both present and future research to capture double standards in judging collective action. In this way, our research situates the distinction between normative and non-normative collective action in the eye of the beholders and opens it to scientific investigation.

The second purpose of our research is to test for double standards in judging collective action---that is, whether people judge the same protest actions as more or less acceptable depending on their own and the protesters' group memberships (Hypothesis 1a, 1b) or on the protesters' cause and how it aligns with their own ideological positions (Hypotheses 2a, 2b). In Experiment 1, we use the instrument developed in Studies 1 and 2 to test for double standards in how acceptable participants consider collective action for workers' rights in the United Kingdom. By varying both the participants' and protesters' group memberships (working/middle class), Experiment 1 compares the relative evidence for Hypotheses 1a and 2a. Extending Experiment 1, Experiment 2 considers both system-challenging and system-defending collective action in the context of the 2020 Black Lives Matter protests in the United States. By varying the protest's cause (for/against defunding the police) as well as the participants' and protesters' group memberships (Black/White), Experiment 2 compares the relative evidence for Hypotheses 1a, 1b, 2a, and 2b. Together, the preregistered experiments provide a comprehensive test of the hypothesized identity- and ideology-based double standards in judging collective action.

@@

<!-- In two studies, we developed an instrument of 25 protest actions to capture double standards in judging collective action. -->

@@

@@

@@

In Experiment 2, we tested five hypotheses. We tested two predictions derived from social identity theory. First, we hypothesize that observers are more likely to judge an action as an acceptable means of protest when the protesters are ingroup members than when they are outgroup members (Hypothesis 1a). Second, we hypothesize that observers are more likely to judge an action as an acceptable means of protest when the goal of the protests would benefit the observers' ingroup (Hypothesis 1b). We will test two predictions derived from system justification theory. First, we hypothesize that observers are more likely to judge an action as an acceptable means of protest when the goal of the protest is to support rather than challenge the system (Hypothesis 2a). Second, we hypothesize that observers who endorse system-justifying beliefs are more likely to consider system-supporting actions acceptable means of protest-and less likely to consider system-challenging actions acceptable means of protest-than observers who do not endorse system-justifying beliefs (Hypothesis 2b). In addition, we will test an alternative hypothesis. We hypothesize that observers are more likely to consider actions as acceptable means of protest when they support the cause of the protesters than when they oppose their cause (Hypothesis 3). We note that Hypotheses 1a, 1b, and 2a make competing predictions about differences between conditions while Hypotheses 2b and 3 need not contradict each other or any of the other hypotheses.

@@

We included additional measures to describe the sample, to describe reactions to the manipulation, or to use in non-preregistered analyses. In addition to demographic questions, we asked participants how outraged they were about recent incidents of police violence against Black Americans, to what extent they identifed with the protesters described in the study, and to what extent they identifed with their racial ingroup (1 = *not at all*, 7 = *very much*). We also asked how often, if at all, participants had participated in protest actions such as the ones we had asked about (1 = *never*, 5 = *very often*) and whether they has participated in protests for reforming, defunding or abolishing the police; against reforming, defunding or abolishing the police; or in neither. We measured political orientation with a one-item semantic differential scale: "People often describe their political orientation as liberal or conservative. On a scale from liberal to conservative, where would you position yourself?"; 1 = *liberal*, 7 = *conservative*).

@@

Figure \ref{fig:f5} shows the predicted pattern of condition-wise differences underlying those fixed effects. Participants who endorsed system-justifying beliefs tended to consider system-challenging and system-defending protest actions to be equally unacceptable. In contrast, participants who rejected system-justifying beliefs evinced ideology-based double standards: In line with Hypothesis 2b, they considered the same protest actions to be more acceptable if the protesters challenged the system or, to a lesser extent, if the protesters were from the disadvantaged group.

@@

We reran Model 3 with political orientation, instead of system justification, as an alternative predictor variable. Our results mirrored results from the preregistered analyses: More liberal participants were more likely to consider protest actions for defunding the police to be acceptable (`r with(m4_exp, b_text[Cause == "for"])`) but more conservative participants were not more likely to consider protest actions against defunding the police to be acceptable (`r with(m4_exp, b_text[Cause == "against"])`). As Figure 5 shows, conservative participants tended to consider all protest actions to be equally unacceptable while liberal participants considered the same protest actions to be more acceptable if protesters rallied around a progressive cause. Our non-preregistered analyses thus found evidence for ideology-based double standards in judging collective action with a different operationalization of political ideology.

@@

- Summary: No evidence for identity-based double standards, but some evidence for ideology-based double standards
- Two plausible interpretations: (1) conservatives oppose controversial protest (no matter the end) *or* (2) the end justifies the mean
- Re (1) it's plausible but what about recent events? (see public polling from grant)
- Re (2) why is there no asymmetry?
- Outline the-end-justifies-the-mean idea, highlighting that just supporting a cause is not enough, it needs moralization
- Discuss why high-status group are more accepting of controversial protest actions
- Discuss limitations (two cultures, two issues, liberal samples; system-defending vs. system-supporting actions; what about other conservative causes?)
- Invite researchers to use our paradigm

>>>>

---
title: "Double Standards in Judging Collective Action"
header: "DOUBLE STANDARDS"
keywords: protest, collective action, double standards, social identity theory, system justification theory, Black Lives Matter
abstract: "Collective action is a powerful force driving social change but often sparks contention about what actions are acceptable means to effect social change. In four studies (total *N* = 2,246), we investigated double standards in judging collective action---that is, whether observers will judge the same protest actions as more acceptable if the protesters are ingroup members and their cause aligns with the ingroup's interests (*identity-based double standards*) or if the protesters' cause aligns with the observers' ideological positions (*ideology-based double standards*). In two studies, we developed an instrument of 25 controversial protest actions, based in item response theory, to measure where people draw the line between acceptable and unacceptable forms of collective action. In two preregistered experiments, we used this instrument to test for double standards in judging collective action for workers' rights in the United Kingdom and for or against defunding the police in the United States. We found evidence for ideology-based, but not for identity-based, double standards: Participants judged the same protest actions to be more acceptable when the protesters' cause aligned with their own (system-challenging) ideological position---but participants showed no ingroup bias when judging collective action by ingroup and outgroup members. Our findings have theoretical and practical implications for understanding the often divided response to prominent social movements."
bibliography: references.bib
csl: apa.csl
output:
  pdf_document:
    latex_engine: pdflatex
    keep_tex: yes
    template: preprint-template.tex

@@

In this paper, we examine double standards in judging collective action---that is, whether observers judge the same protest actions as more or less acceptable depending on their own and the protesters' group memberships (*identity-based double standards*) or on the protesters' cause and how it aligns with their own ideological positions (*ideology-based double standards*). In two studies, we first develop an instrument of 25 controversial protest actions, based in item response theory, to capture double standards in judging collective action. In two preregistered experiments, we then use this instrument to test for double standards in judging protest actions for workers' rights in the United Kingdom and for or against defunding the police in the United States. We find evidence for ideology-based, but not for identity-based, double standards: People judge the same protest actions to be more acceptable when the protesters' cause aligns with their own (system-challenging) ideological positions---but do not show ingroup bias when judging collective action by ingroup and outgroup members.

<!-- In the next sections, we, first, introduce the consequential distinction between normative and nonnormative collective action and define double standards in judging what actions are acceptable and unacceptable means to effect social change. We, second, derive predictions from social psychological theories about identity- and ideology-based double standards that might affect judgements about collective action. We, third, introduce item response theory as a conceptual and statistical framework for study double standards in judging collective action. We, finally, provide an overview of the present research. -->

@@

Collective action can take many forms. Psychologists distinguish between *normative* collective action that seeks to achieve a political goal while conforming to the norms of the existing social system and *non-normative* collective action that violates those norms [@wright_responding_1990]. Recent research has shown that this distinction matters for how people respond to collective action. @shuman_disrupting_2021 showed that non-normative, non-violent collective action is more effective than either normative or violent collective action at gaining concessions from those most resistant to social change. @feinberg_activists_2020 demonstrated, however, that observers are less supportive of social movements that use extreme protest actions. Similarly, @teixeira_is_2020 found that advantaged-group members perceive non-normative collective action by disadvantaged-group members as more damaging to their ingroup's social image and are thus less supportive of such action. @verkuyten_intolerance_2022 found that, in general, the public tends to be intolerant of non-normative collective action. This means that activists face the dilemma that non-normative collective action might be most effective at gaining concessions but that it also risks reducing popular support for a cause. Together, these studies show that the distinction between normative and non-normative collective action is psychologically and societally consequential.

@@

Just as system justification shapes support for, and opposition to, collective action, we propose that it results in ideology-based double standards in what actions people consider acceptable forms of collective action. First, we hypothesize that  people will judge the same protest actions as more acceptable if the protest supports, rather than challenges, the system (*Hypothesis 2a*). This hypothesis reflects the assumption that the motivation to justify the system is, to some extent, universal. Second, we hypothesize that people will judge the same protest actions as more acceptable if they endorse system-justifying beliefs and the protest supports the system or if they reject system-justifying beliefs and the protest challenges the system (*Hypothesis 2b*). Notably, this hypothesis implies a symmetry in how people with different ideological orientations judge system-supporting and system-challenging collective action.

@@

In this research, item response theory serves two purposes. First, item response theory provides a statistical framework for scale development. In Studies 1 and 2, we select the most discriminating and informative actions to build an instrument for capturing double standards in the experiments. Second, item response theory provides a formal and statistical definition of double standards in judging collective action as respondents being more accepting of the same protest actions depending on who the protesters and what they are protesting. In Experiments 1 and 2, we operationalize double standards as the differences in $\theta_{(z)}$ between conditions which, as $\theta_{(z)}$ is $z$-standardized, correspond to Cohen's $d$ effect sizes.

@@

The first purpose of our research is to develop an instrument to measure where people draw the line between acceptable and unacceptable forms of collective action. In Study 1, we ask participants to generate more and less extreme protest actions and compile a pool of protest actions for scale development. In Study 2, we ask participants to rate how acceptable they consider each of those protest action to be and use item response theory to select the most discriminating and informative protest actions for our instrument. In so doing, we build an instrument that enables both present and future research to capture double standards in judging collective action. In this way, our research situates the distinction between normative and non-normative collective action in the eye of the beholders and opens it to scientific investigation.

The second purpose of our research is to test for double standards in judging collective action---that is, whether people judge the same protest actions as more acceptable if the protesters' are ingroup members or the protesters' cause aligns with the ingroup's interests (Hypothesis 1a, 1b) or if the protesters' cause aligns with their own system-justifying or system-challenging ideological orientation (Hypotheses 2a, 2b). In Experiment 1, we use the instrument developed in Studies 1 and 2 to test for double standards in how acceptable participants consider collective action for workers' rights in the United Kingdom. By varying both the participants' and the protesters' group memberships (working/middle class), Experiment 1 compares the relative evidence for identity-based (Hypotheses 1a, 1b) and ideology-based double standards (Hypothesis 2a). Extending Experiment 1, Experiment 2 considers both system-challenging and system-defending collective action in the context of the 2020 Black Lives Matter protests in the United States. By varying the protesters' cause (for/against defunding the police) as well as the participants' and the protesters' group memberships (Black/White), Experiment 2 compares the relative evidence identity-based (Hypotheses 1a, 1b) and ideology-based double standards (Hypothesis 2a, 2b). Together, the preregistered experiments provide a comprehensive test of the hypothesized double standards in judging collective action.

@@

In two studies, we developed an instrument of 25 protest actions to capture double standards in judging collective action.

@@

In Experiment 1, we used the newly developed instrument to test for double standards in judging collective action for workers' rights in the United Kingdom. That is, we tested whether people with either working-class or professional jobs applied different standards when judging collective action taken by people with either working-class or middle-class jobs to protest against a fictitious government bill threatening their rights.

By varying the participants' and the protesters' group memberships, we tested the preregistered hypothesis that, in line with Hypothesis 1a and 1b, participants would judge the same protest actions as more acceptable if the protesters were ingroup members protesting for the ingroup's rights than if they were outgroup members protesting for the outgroup's rights. By comparing reactions to protesters with working-class or professional jobs, we tested how observers judge collective action by lower-status (i.e. working-class) and higher-status (professional/middle-class) group members. Assuming that collective action by lower-status group members is more threatening to the unequal status than collective action by higher-status group members, we tested the preregistered hypothesis that, in line with Hypothesis 2a, participants would judge the same protest actions as more acceptable if the protesters were from a higher-status group than if they were from a lower-status group. In non-preregistered analyses, we further explored whether, in line with Hypothesis 2b, participants would judge the same protest actions as more or less acceptable depending on their own ideological positions. In this way, Experiment 1 provided a first test of the hypothesized identity- and ideology-based double standards in judging collective action.

@@

In Experiment 2, we investigated potential double standards in judging collective action for or against defunding the police in the United States. That is, we tested whether Black and White Americans applied different standards when judging collective action taken by either Black or White protesters to protest either for or against police divestment as a possible solution to racist police violence.

@@

<!-- In Experiment 2, we tested five hypotheses. We tested two predictions derived from social identity theory. First, we hypothesize that observers are more likely to judge an action as an acceptable means of protest when the protesters are ingroup members than when they are outgroup members (Hypothesis 1a). Second, we hypothesize that observers are more likely to judge an action as an acceptable means of protest when the goal of the protests would benefit the observers' ingroup (Hypothesis 1b). We will test two predictions derived from system justification theory. First, we hypothesize that observers are more likely to judge an action as an acceptable means of protest when the goal of the protest is to support rather than challenge the system (Hypothesis 2a). Second, we hypothesize that observers who endorse system-justifying beliefs are more likely to consider system-supporting actions acceptable means of protest-and less likely to consider system-challenging actions acceptable means of protest-than observers who do not endorse system-justifying beliefs (Hypothesis 2b). In addition, we will test an alternative hypothesis. We hypothesize that observers are more likely to consider actions as acceptable means of protest when they support the cause of the protesters than when they oppose their cause (Hypothesis 3). We note that Hypotheses 1a, 1b, and 2a make competing predictions about differences between conditions while Hypotheses 2b and 3 need not contradict each other or any of the other hypotheses. -->

@@

We included additional measures to describe the sample, to describe reactions to the manipulation, or to use in non-preregistered analyses. In addition to demographic questions, we asked participants how outraged they were about recent incidents of police violence against Black Americans, to what extent they identifed with the protesters described in the study, and to what extent they identifed with their racial ingroup (1 = *not at all*, 7 = *very much*). We also asked how often, if at all, participants had participated in protest actions such as the ones we had asked about (1 = *never*, 5 = *very often*) and whether they has participated in protests for reforming, defunding or abolishing the police; against reforming, defunding or abolishing the police; or in neither. We measured political orientation with a one-item semantic differential scale: "People often describe their political orientation as liberal or conservative. On a scale from liberal to conservative, where would you position yourself?" (1 = *liberal*, 7 = *conservative*).

@@

Figure \ref{fig:f5} shows the estimated pattern of condition-wise differences underlying those fixed effects. Participants who endorsed system-justifying beliefs tended to consider system-challenging and system-defending protest actions to be equally unacceptable. In contrast, participants who rejected system-justifying beliefs evinced ideology-based double standards: In line with Hypothesis 2b, they considered the same protest actions to be more acceptable when the protesters challenged the system or, to a lesser extent, when the protesters were from the disadvantaged group.

@@

We reran Model 2 with political orientation, instead of system justification, as an alternative predictor variable. Our results mirrored results from the preregistered analyses: More liberal participants were more likely to consider protest actions for defunding the police to be acceptable (`r with(m4_exp, b_text[Cause == "for"])`) but more conservative participants were not more likely to consider protest actions against defunding the police to be acceptable (`r with(m4_exp, b_text[Cause == "against"])`). As Figure 5 shows, conservative participants tended to consider all protest actions to be equally unacceptable while liberal participants considered the same protest actions to be more acceptable when protesters rallied around a progressive cause. Our non-preregistered analyses thus replicated the ideology-based double standards from the preregistered analyses with a different operationalization of political ideology.

@@

<!-- Our research was motivated by examples showing that different groups judge the same protest action more acceptable depending on who's doing the action. That is, in contrast to previous research, not the case. -->

- Summary: No evidence for identity-based double standards, but some evidence for ideology-based double standards
- Reject the alternative hypothesis that there is consensus about what is an acceptable means of protest.
- Relate our findings to @verkuyten_intolerance_2022 
- Two plausible interpretations: (1) conservatives oppose controversial protest (no matter the end) *or* (2) the end justifies the mean
- Re (1) it's plausible but what about recent events? (see public polling from grant)
- Re (2) why is there no asymmetry?
- Outline the-end-justifies-the-mean idea, highlighting that just supporting a cause is not enough, it needs moralization
- Discuss why high-status group are more accepting of controversial protest actions
- Discuss limitations (two cultures, two issues, liberal samples; system-defending vs. system-supporting actions; what about other conservative causes?; ingroup bias might need zero-sum situation to succeed)
- Invite researchers to use our paradigm
